---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Read the activity dataset.
```{r echo= TRUE}
df<- read.csv('activity/activity.csv', stringsAsFactors = FALSE)
head(df)
```

Preprocess the data to set columns to proper datatypes, for further analysis.

```{r echo= TRUE}
df$steps<- as.integer(df$steps)
df$date<- as.Date(df$date, format = "%Y-%m-%d")
df$interval<- as.integer(df$interval)
head(df)
```

## What is mean total number of steps taken per day?
First we find the total number of steps (sum) taken per day.
```{r echo=TRUE}
df1<-aggregate(df$steps, by = list(df$date), FUN = sum, na.rm =TRUE)
names(df1)<- c('Date', 'Total_Steps_PerDay')
head(df1, 20)
```

Show a histogram for total number of steps taken each day. 'Green' vertical line marks the mean, whereas the 'Red' vertical line marks the median.
```{r echo=TRUE}
hist(df1$Total_Steps_PerDay, main = 'Ignoring missing data: Histogram for Total Steps per day', xlab = 'Daily Steps')
abline(v = mean(df1$Total_Steps_PerDay), col = 'green', lwd =3)
abline(v = median(df1$Total_Steps_PerDay), col = 'red', lwd =3)
```

Calculate the mean total steps taken per day - 
```{r echo=TRUE}
print(paste('mean:', round(mean(df1$Total_Steps_PerDay),0), 'steps'))
print(paste('median:', median(df1$Total_Steps_PerDay), 'steps'))
```


## What is the average daily activity pattern?
```{r echo=TRUE}
df3<- aggregate(df$steps, by = list(df$interval), FUN = mean, na.rm = TRUE)
names(df3)<- c('Interval (mins)', 'Average_Steps_PerInterval')
df3$Average_Steps_PerInterval<- round(df3$Average_Steps_PerInterval, 2)
head(df3, 20)
```

A plot of the daily activity pattern - 
```{r echo=TRUE, fig.width= 12, fig.height= 4}
plot(x = df3$`Interval (mins)`, y = df3$Average_Steps_PerInterval, type = 'l', main = "Daily activity pattern", xlab = "Time interval (min)", ylab = "Average number of steps")
```
Maximum steps interval is -
```{r echo=TRUE}
paste(df3[which.max(df3$Average_Steps_PerInterval), 'Interval (mins)'], "mins")
```



## Imputing missing values
```{r echo=TRUE}
print(paste('Total missing values:' , sum(is.na(df$steps))))
```


We fill na's by the average steps in the same interval.
```{r echo=TRUE}
df4<- df
for(i in 1:nrow(df4)){
  if(is.na(df4[i, 'steps']))
  df4[i, 'steps']<- df3[df3$`Interval (mins)` == df4$interval[i], 'Average_Steps_PerInterval']
}
head(df4, 20)
```

Plot the histogram of dataset after missing values have been imputed. Calculate the new mean and median.
```{r echo=TRUE}
df5<-aggregate(df4$steps, by = list(df4$date), FUN = sum, na.rm =TRUE)
names(df5)<- c('Date', 'Total_Steps_PerDay')
hist(df5$Total_Steps_PerDay, main = 'After treating missing values: Histogram for Total Steps per day', xlab = 'Daily Steps')
```
Calculate the mean total steps taken per day after imputing the missing values - 
```{r echo=TRUE}
print(paste('new mean:', round(mean(df5$Total_Steps_PerDay),0), 'steps'))
print(paste('new median:', median(df5$Total_Steps_PerDay), 'steps'))
```
This method of missing data imputation increases the mean and median of the data.

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE, fig.width= 12, fig.height= 5, warning=FALSE}
library(ggplot2)
library(ggpubr)

df4$Weekday<- weekdays(df4$date, abbreviate = T)
df4$Weekend<- ifelse(df4$Weekday %in% c('Sat', 'Sun'), 'weekend', 'weekday')

#weekends
df6<- aggregate(df4[df4$Weekend == "weekend",]$steps, by = list(df4[df4$Weekend == "weekend",]$interval), FUN = mean, na.rm = TRUE)
names(df6)<- c('Interval (mins)', 'Average_Steps_PerInterval')
df6$Average_Steps_PerInterval<- round(df6$Average_Steps_PerInterval, 2)
df6$Weekend<- "weekend"

#weekdays
df7<- aggregate(df4[df4$Weekend == "weekday",]$steps, by = list(df4[df4$Weekend == "weekday",]$interval), FUN = mean, na.rm = TRUE)
names(df7)<- c('Interval (mins)', 'Average_Steps_PerInterval')
df7$Average_Steps_PerInterval<- round(df7$Average_Steps_PerInterval, 2)
df7$Weekend<- "weekday"

p1<- ggplot(df6, aes(y = Average_Steps_PerInterval, x = `Interval (mins)`))+geom_line(size =2, color = "blue")+ labs( x = NULL, y = NULL, title = NULL) + facet_grid(.~Weekend)+theme_light()+ theme(strip.background.x = element_rect(fill="#FFFFE0"),strip.text = element_text(size=12, colour="#000000", hjust = 0.5, vjust = 0.5), axis.ticks.x = element_blank(), axis.text.x = element_blank())
p2<- ggplot(df7, aes(y = Average_Steps_PerInterval, x = `Interval (mins)`))+geom_line(size =2, color = "blue")+ labs( x = NULL, y = NULL, title = NULL)+ facet_grid(.~Weekend) + theme_light()+ theme(strip.background.x = element_rect(fill="#FFFFE0"),strip.text = element_text(size=12, colour="#000000", hjust = 0.5, vjust = 0.5))

p<- ggarrange(p1, p2, ncol = 1)
annotate_figure(p , bottom = "Interval (min)", left = text_grob("Number of steps", rot = 90))
```

